---
title: "Final Report"
author: "Sergio O."
date: "April 11, 2025"
bibliography: references.bib
csl: ecology.csl
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Introduction

Delete this placeholder text, and replace it with your introduction text.

# Methods

The data I used was collected by NEON technicians [@thorpe2016introduction]. I will be working with NEON small mammal data [@parsons2023monitoring].

## Site Locations and Descriptions

Add text here.

## Field Sampling Design

Add text here.

## Data Analysis and Visualization

I used the ggplot2 package [@ggplot2] in R to do my visualizations.

# Results

## Subsections are ok in the results section too

Add a number of code chunks in the Results section. These should read in, subset and plot the data as needed (no need to save any figures to pdf, since they will be put into the rendered document when you click 'knit'), and, for any hypotheses that you want to test, an appropriate statistical test. 

```{r load-packages}
# If you add any additional packages here, make sure they are
# also listed in the DESCRIPTION file
library("dplyr")
library("tidyr")
library("ggplot2")
library("readr")
library("lubridate")

# load biomass data
write.csv(megapit_cn, "data/raw_data/megapit_carbon_nitrogen.csv", row.names = FALSE)
```




# Discussion

# Figures

```{r Nitrogen-%-Distrubution}

# Plot Time! Nitrogen % Distrubution 

# Plot 1: Nitrogen Distribution
p1 <- ggplot(megapit_cn, aes(x = nitrogenPercent)) +
  geom_histogram(binwidth = 0.2, fill = "#4CAF13", color = "black") +
  labs(title = "Distribution of Nitrogen Content in Roots",
       x = "Nitrogen Percent",
       y = "Number of Samples") +
  theme_minimal()
print(p1)



# Plot 2: Carbon vs Nitrogen
p2 <- ggplot(megapit_cn, aes(x = carbonPercent, y = nitrogenPercent)) +
  geom_point(alpha = 0.6, color = "#FF9800") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Carbon vs Nitrogen in Megapit Root Samples",
       x = "Carbon Percent",
       y = "Nitrogen Percent") +
  theme_minimal()
print(p2)



# Plot 3: C:N Ratio
megapit_cn$cn_ratio <- megapit_cn$carbonPercent / megapit_cn$nitrogenPercent
p3 <- ggplot(megapit_cn, aes(x = cn_ratio)) +
  geom_histogram(binwidth = 2, fill = "#9C27B0", color = "black") +
  labs(title = "Carbon:Nitrogen Ratio Distribution",
       x = "C:N Ratio",
       y = "Count") +
  theme_minimal()
print(p3)




# Plot for Megapit root carbon-nitrogen consumption as Nitrogen Vs Carbon

# Step 1: Subset and join
carbon_nitrogen <- megapit_data[["mpr_carbonNitrogen"]]
root_sample <- megapit_data[["mpr_perrootsample"]]

joined_data <- carbon_nitrogen %>%
  left_join(root_sample, by = join_by(cnSampleID == sampleID))

# Step 2: Filter usable rows
filtered_data <- joined_data %>%
  filter(!is.na(sizeCategory),
         !is.na(nitrogenPercent),
         !is.na(carbonPercent))

# Step 3: Pivot to long format
long_data <- filtered_data %>%
  pivot_longer(cols = c(nitrogenPercent, carbonPercent),
               names_to = "element",
               values_to = "percent")

# Step 4: Summarize mean + standard error for error bars
summary_data <- long_data %>%
  group_by(sizeCategory, element) %>%
  summarise(mean_percent = mean(percent),
            se = sd(percent) / sqrt(n()), .groups = "drop")

# Step 5: Create bar plot with error bars and data labels
ggplot(summary_data, aes(x = sizeCategory, y = mean_percent, fill = element)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = mean_percent - se, ymax = mean_percent + se),
                position = position_dodge(width = 0.9),
                width = 0.2) +
  geom_text(aes(label = round(mean_percent, 1)),
            position = position_dodge(width = 0.9),
            vjust = -1.2, size = 4) +  # 🔧 Raised above bars for visibility
  scale_fill_manual(values = c("nitrogenPercent" = "red",
                               "carbonPercent" = "blue"),
                    labels = c("Nitrogen", "Carbon")) +
  labs(title = "Nitrogen vs Carbon Consumption by Root Size Category",
       x = "Root Size Category",
       y = "Mean Percent (%)",
       fill = "Element") +
  theme_minimal() +
  theme(legend.position = "top") +
  expand_limits(y = max(summary_data$mean_percent + summary_data$se) + 5)  # 📈 Extra headroom for labels





                                   
# Joining Data Example 

# Subset out the relevant data frames
carbon_nitrogen <- megapit_data[["mpr_carbonNitrogen"]]
root_sample <- megapit_data[["mpr_perrootsample"]]

# Check actual join column names
# carbon_nitrogen uses: "cnSampleID"
# root_sample uses: "sampleID"

# Perform the join safely
joined_data <- carbon_nitrogen %>%
  left_join(root_sample, by = join_by(cnSampleID == sampleID))

# Filter and remove any NA values in key columns
high_nitrogen_samples <- joined_data %>%
  filter(!is.na(nitrogenPercent),
         !is.na(carbonPercent),
         nitrogenPercent > 0.5)  # lowered threshold for more visible data

# Plot the data
ggplot(high_nitrogen_samples, aes(x = carbonPercent, y = nitrogenPercent)) +
  geom_point(color = "blue") +
  labs(title = "Samples with Nitrogen > 0.5% (Joined Tables)",
       x = "Carbon Percent",
       y = "Nitrogen Percent") +
  theme_minimal()


  
```

**Figure 1:** *Fontinalis neomexicana* seems to have higher dry mass than other taxa in this site.

# Sources Cited

